Non-parametric approaches to optimal policy are more robust

### 2013-03-27 inst/doc/nonparametric-bayes.Rmd
updates to pomp during our discussion today
writing writing

Most management recommendations from the ecological literature are based on
(or at least motivated by) parametric models. Though in principle these models
reflect mechanistic underpinnings of biological interactions involved, in practice
real mechanisms are often unknown and the true dynamics too complex to be captured
by simple models [@refs]. While simple mechanistic models can nevertheless
provide imporant insights into possible dynamics -- for instance, demonstrating
that a vaccine does not have to be 100% effective to eliminate the transmission of a virus [@Kermack1921] -- such approaches are not well-suited for use in 
forecasting future dynamics upon which outcomes management policy can be based.  
Despite a long history of work emphasizing the difference between modeling 
for understanding (generic or strategic models) and modeling for prediction (precise, specific, or tactical models) [e.g. @Levins1966], simple parametric models 
continue to find application in predictive management contexts.  The management
of marine fisheries is a prime example, in which simple, mechanistically 
motivated models of stock-recruitment relationships such as Ricker or Beverton-Holt
curves are systematically fit to data and used in solving decision-theoretic
or optimal control problems determining resource management policies such
as maximum sustainable yield [@refs].

<!-- Better defintion of focal problems here .... -->
Stochastic dynamic programming @Mangel1982

To address structural uncertainty in the processes involved, such mechanistic models
can be adapted to accomidate uncertainty in the parameters or consider a set of 
alternative models simultaneously (belief SDP)

Structural uncertainty.

Further unknowns such as measurement uncertainty, parameter uncertainty,
unobserved states, knowledge of boundary conditions, etc. further
compound the issue.  Though a hierarchical Bayesian approach provides a
natural way to address these from a statistical standpoint, formulating
reasonable parametric descriptions of each form of uncertainty is a
challenging task in itself, let alone the computational difficulties
of solving such a system. @Cressie2009 provides a good account of the
successess and challenges of the approach.  Applying these approaches in
the management context of sequential decision making, in which forecasts
must be obtained over a range of possible actions and updated regularly
as new information arrives makes such an approach less feasible still.

Though machine learning approaches have begun to appear in the
ecological and conservation literature (Species distribution models),
including the Gaussian process based approach used here [@Munch2005],
they remain unfamiliar and untrusted approaches for most ecologists.
Machine learning approaches represent an essentially pattern-based
rather than process-based approach, raising the same skepticism from most
theoretical ecologists that they hold for older correlative methods such
as linear regression, while the complexity of the techniques has barred
their adoption in more empirical audiences. <!-- Whoa! rampant speculation
alert! -->

<!-- Whoa, isn't machine learning used to make decisions all the time,
 eg, in engineering and Artificial Intelligence?  -->

A more immediate barrier to their adoption is the absence of a framework
for applying machine learning approaches to resource management problems.
Traditional approaches to optimal control (Pontryagin's principle, stochastic
dynamic programming) rely on knowledge of the state equation, usually described
by a simple parametric model. Here we illustrate how a stochastic dynamic
programming algorithm can alternatively be driven by the predictions from
a Gaussian process -- a machine learning approximation to the state dynamics.  
<!-- Okay, perhaps that's novel, but it's pretty trivial.  Isn't it
obvious to everyon that it's trivial? -->

@@

>>>>

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological interactions
involved, in practice real mechanisms are often unknown and the true
dynamics too complex to be captured by simple models [@refs]. While
simple mechanistic models can nevertheless provide imporant insights
into possible dynamics -- for instance, demonstrating that a vaccine
does not have to be 100% effective to eliminate the transmission of a
virus [@Kermack1921] -- such approaches are not well-suited for use in
forecasting outcomes of potential management options.  Non-parametric 
approaches offer a more flexible alternative that can both more accurately
reflect the data available while also representing greater uncertainty
in areas (of state-space) where data is lacking.  Ecological research
and management strategy should pay closer attention to the opportunities 
and challenges nonparametric modeling can offer.

### Quantitative vs Qualitative Decisions

In this paper, we consider those ecological management problems in which
a mathematical (or computational) model is used to quantitatively inform
decision-making by allowing a manager to compare to the expected consequences
of potential management actions (or policies).  We distinguish this from the
solely qualitative use of a model, in which models are used to represent and
compare hypotheses of different mechanisms that would lead to qualitatively 
different actions.

We argue that while simple mechanistically motivated models
may be best for the latter case [@Cunningham2013; @Geritz2012], such models 
can be not only inaccurate but misleading in quantitative decision making.
Nonparametric models can more realistically represent uncertainties outside 
of observed data while also better capturing the dynamics in the region
observed.  Advances in the theory and computational implementations of 
nonparametric methods make them ripe for such applications.  We use the classic
problem of optimal harvest of a marine fishery to illustrate how the 
nonparametric approach of Gaussian processes can be applied.

We imagine the manager seeks to maximize some known socio-economic value function, 
given their knowledge of the system. 
In particular, we will focus on the case in which the manager faces a series of decisions,
such as setting a harvest quota for a marine fishery each year based on observations
of the stock size the year before.

### Uncertainties: the classical approach

The ecological dynamics of most systems of management interest are 
typically both complex and unknown [@refs].  Despite this, quantitative policy
optimization in these systems has almost exclusively been based on
simple, mechanistically motivated parametric models.
There are several reasons for this.  First, limited available data almost 
always percludes any accurate estimate of more complicated models.  Second,
the computational demands of determining an optimal stategy in a sequential
decision-theory problem suffer from the "curse of dimensionality" that makes  
it difficult to consider more complicated models, states, or actions.  Unfortunately,
these challenges also make it particularly difficult to reflect the true uncertainty
either in the parameters of the given models (parametric uncertainty) or in 
the structure of the models themselves (structural uncertainty).  
Further unknowns such as measurement uncertainty, parameter uncertainty,
unobserved states, knowledge of boundary conditions, etc. further
compound the issue.  Though a hierarchical Bayesian approach provides a
natural way to address these from a statistical standpoint, formulating
reasonable parametric  descriptions of each form of uncertainty is a
challenging task in itself, let alone the computational difficulties
of solving such a system. @Cressie2009 provides a good account of the
successess and challenges of the approach.  Applying these approaches in
the management context of sequential decision making, in which forecasts
must be obtained over a range of possible actions and updated regularly
as new information arrives makes such an approach less feasible still.

An active literature and growing computational power over the past several decades have
only marginally improved this situation.  Parametric or structural uncertainty
can be introduced only by increasing the state space to reflect a distribution
of possible parameters [@refs] or the degree of belief in each of a set of possible models [@refs]
Frequently, the space of possible actions must then be reduced or the algorithms adjusted
by approximations to keep computations feasible.

Though machine learning approaches have begun to appear in the
ecological and conservation literature (Species distribution models),
including the Gaussian process based approach used here [@Munch2005],
they remain unfamiliar and untrusted approaches for most ecologists.
One potential barrier to their adoption is the absence of a framework
for applying machine learning approaches to resource management problems.
Traditional approaches to optimal control (Pontryagin's principle, stochastic
dynamic programming) rely on knowledge of the state equation, usually described
by a simple parametric model. Here we illustrate how a stochastic dynamic
programming algorithm can alternatively be driven by the predictions from
a Gaussian process -- a machine learning approximation to the state dynamics.

<!-- Okay, perhaps that's novel, but it's pretty trivial.  Isn't it
obvious to everyon that it's trivial? -->

@@

### The fisheries management problem

In our example, we focus on the problem in which a manager must set 
the harvest level for a marine fishery each year to maximize the net
present value of the resource, given an estimated stock size from the year
before. Rich data and global concerns have made marine fisheries the crucible for much
of the founding work [@Gordon1954; @Reed1979; @May1979; @Ludwig1982]
in managing ecosystems under uncertainty.  Global trends [@Worm2006]
and controversy [@Hilborn2007; @Worm2009] have made understanding these
challenges all the more pressing.

To permit comparisons against a theoretical optimum we will consider
data on the stock dynamics simulated from a simple parametric model
in which recruitment of the fish stock $X_{t+1}$ in the following year
is a stochastic process governed by a function $f$ of the current 
stock $X_t$, selected harvest policy $h_t$, and noise process $Z$,

$$X_{t+1} = Z_t f(X_t, h_t) $$

Given parameters for the function $f$ and probability distribution $Z$, 
along with a given economic model determining the  price/profit $\Pi(X_t, h_t)$
realized in a given year given a choice of harvest $h_t$ and observed stock $X_t$.
This problem can be solved exactly for discretized values of stock $X$ and policy $h$
using stochastic dynamic programming [@Mangel1982].  
Problems of this sort underpin much marine fisheries management today.

A crux of this approach is correctly specifying the functional form of $f$,
along with its parameters.  The standard approach [@refs] uses one of a 
handful of common parametric models representing the stock-recruitment
relationship, usually after estimating the model parameters from any 
available existing data. Though modifications of this approach mentioned in the 
introduction can permit additional sources of uncertainty such as measurement 
error in the stock assessment, implementation errors in the harvest policy, [@Clark1986, @Roughgarden1996, @Sethi2005]
uncertainty in parameters [@Mangel1985, @Schaupaugh2013] or model structure [@Williams2001, @Athanassoglou2012]

Here, we compare this approach to our alternative
that uses a Gaussian Process in place of the a given stock recruitment curve.

### The Non-parametric Bayesian alternative for stock-recruitment curves

Following [@Munch2005], we will estimate Gaussian processes from

###
